Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-31 Manuscript/manuscriptPSsuppl.tex
Updates based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21 and some other updates

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of other lipids with bound charge,
like negatively charged PS and PG lipids,  are also systematic,
but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol and cationic surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

@@

\pagebreak
\section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
Before using the headgroup order parameters to compare ion binding affinity between simulations
and experiments, the response of the order parameters to the bound charge in simulations should be quantified
against experiments~\cite{catte16,melcr18}. In our previous work~\cite{catte16},
the ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ was in good agreement with the experiments~\cite{akutsu81}
in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
the headgroup order parameter responses were compared more carefully with the experiments of cationic
surfactants in POPC bilayer~\cite{scherer89}. The comparison reveals that the both headgroup
order parameters in the Lipid14 model are too sensitive to the bound charge,
while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
Therefore, the headgroup order parameter response to the bound charge is actually more realistic
in CHARMM36 model than in the Lipid14. In the latter, both order parameters are equally too 
sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.
The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
is relatively more sensitive than the $\alpha$-carbon order parameter. These results have to be taken
into account when analysing the ion binding affinities using headgroup order parameters in simulations.
However, we conclude that the discrepancies arising from the sentivity of lipid headgroup to
bound charge are typically smaller than then discrepancies arising from ion binding affinity.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer from simulations and experiments \cite{scherer89}.
  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
}
\end{figure}

\pagebreak
\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
  \caption{\label{concentrationDEFfigure}
    Difference between order parameter changes as a function of CaCl$_2$ concentration demonstrated
    for the two possible definitions of ion concentration, concentration in buffer before solvating the
    lipids (buffer concentration) and concenrtration in bulk after solvating the lipids (bulk concentration) \cite{melcr18}. 
}
\end{figure}

@@

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 have be written based on issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}

%%
%% I think that this section is not very useful, I think that we should remove it.
%%
%\section{Sodium binding to DMPC:DOPS mixture}
%
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
%  \caption{\label{PSresponseTONaClDMPC}
%    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
%  }
%\end{figure}
%The experimental results show essentially no changes in the order parameters as a function of
%added NaCl, while significant changes are observed in simulations. However,
%the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
%Therefore, we cannot exclude the possibility that the NaCl induced changes were already
%saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
\pagebreak

>>>>

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of other lipids with bound charge,
like negatively charged PS and PG lipids,  are also systematic,
but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

@@

\pagebreak
\section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
Before using the headgroup order parameters to compare ion binding affinity between simulations
and experiments, the response of the order parameters to the bound charge in simulations should be quantified
against experiments~\cite{catte16,melcr18}. In our previous work~\cite{catte16},
the ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ was in good agreement with the experiments~\cite{akutsu81}
in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
the headgroup order parameter responses were compared more carefully with the experiments of
cationic dihexadecyldimethylammonium bromide (DHAB) surfactants in POPC bilayer~\cite{scherer89}.
The comparison reveals that the both headgroup order parameters in the Lipid14 model are too sensitive to the bound charge,
while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
Therefore, the headgroup order parameter response to the bound charge is actually more realistic
in CHARMM36 model than in the Lipid14. In the latter, both order parameters are equally too 
sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.
The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
is relatively more sensitive than the $\alpha$-carbon order parameter. These results have to be taken
into account when analysing the ion binding affinities using headgroup order parameters in simulations.
However, we conclude that the discrepancies arising from the sentivity of lipid headgroup to
bound charge are typically smaller than then discrepancies arising from ion binding affinity.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer from simulations and experiments \cite{scherer89}.
  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
}
\end{figure}

\pagebreak
\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
Previous studies using the electrometer concept to assess the ion binding affinity to lipid
bilayers report ion concentrations either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. The difference between these two concentrations increases
with the increasing ion binding affinity. However, the difference is essentially negligible
in the recent model \cite{melcr18} with realistic ion binding affinity (Fig. \ref{concentrationDEFfigure}).
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
  \caption{\label{concentrationDEFfigure}
    Difference between order parameter changes as a function of CaCl$_2$ concentration demonstrated
    for the two possible definitions of ion concentration, concentration in buffer before solvating the
    lipids (buffer concentration) and concentration in bulk after solvating the lipids (bulk concentration),
    from the recent model with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}. 
}
\end{figure}

@@

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameters of POPC headgroup in the mixture are available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}.

In simulations with Berger and CHARMM36 models, the responses of both headgroup order parameter
of POPC and the $\beta$ carbon of POPS to the added sodium or potassium are similar to the experiments
of LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of the counterions 
in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
(Fig. \ref{CIdensPSOCmixt}), but significantly larger error bars and
less systematic changes in the order parameters (Fig. \ref{PSresponseTONaCl}).
Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
for which the data is not shown due to the formation of
ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
Such clusters appering also in MagRog simulations with 4 M concentration of KCl
could explain the unsystematic changes of the order parameters in this model with the added KCl.
In conclusion, the results are in line with the section \ref{ciBINDINGsection}
in the main text, suggesting that the MacRog simulations with KCl give the most
realistic surface charge at the lipid bilayer interface.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1 M of NaCl (left)
    and MacRog simulations with 4 M of KCl (right).
\label{ionCLUSTERS}
  }
\end{figure}
\newpage

%%
%% I think that this section is not very useful, I think that we should remove it.
%%
%\section{Sodium binding to DMPC:DOPS mixture}
%
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
%  \caption{\label{PSresponseTONaClDMPC}
%    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
%  }
%\end{figure}
%The experimental results show essentially no changes in the order parameters as a function of
%added NaCl, while significant changes are observed in simulations. However,
%the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
%Therefore, we cannot exclude the possibility that the NaCl induced changes were already
%saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
